---
title: "HLA typing / Covid"
subtitle: "Report (3rd version, mk3)"
author: "January Weiner"
date: "`r Sys.Date()`"
outputdupa: powerpoint_presentation
always_allow_html: true
outputpipa: 
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
    toc_depth: 3
    fig_caption: yes
    number_sections: true
output: word_document
outputcipa:
  ioslides_presentation: 
    widescreen: true
    smaller: true
    css: files/style.css
    logo: files/logo_bihealth_en.png
toc: no
---

```{r}
patient_IDs_to_exclude <- c(
"C19-CB-0092", # cc90
"C19-CB-0022", # cc20
"C19-CB-0051", # cc49
"C19-CB-0120", # cc117
"C19-CB-0067", # cc65
"C19-CB-0003", # cc3
"C19-CB-0005", # cc4
"C19-CB-0060", # cc58
"C19-CB-0063", # cc61
"C19-CB-0023"  # cc21
)
```

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
```

```{r libraries,cache=FALSE}
library(tidyverse)
library(XML)
library(colorDF)
library(readxl)
library(writexl)
library(cowplot)
library(splitstackshape)
library(HIBAG)
library(pander)
library(beeswarm)
library(randomForest)
library(FactoMineR)
library(factoextra)
library(DT)
source("functions.R")
```

```{r parameters}
min_freq    <- 0.05
min_patient <- 5
```

```{r read_files}
message("reading files")

clin1 <- read_xlsx("meta_data/Copy of clinicaldata_cdna_PS_13.10.2020-Updated.xlsx")
clin2 <- read_xlsx("meta_data/Copy of clinicaldata_cdna_PS_13.10.2020-Updated.xlsx", sheet=2)
cc <- intersect(colnames(clin1), colnames(clin2))

clin <- clin_orig <- rbind(clin1[, cc], clin2[, cc]) %>% 
  rename(ID=1, intub="Intubation (Yes=1/No=0)", death="Tod (Yes=1/No=0)", icu="ICU Y/N") %>% 
  mutate(ID=tolower(ID), icu=toupper(gsub(" .*", "", icu))) %>%
  filter(!is.na(ID))

clin$troponine     <- as.numeric(gsub(" .*", "", clin[["Troponin -T [ng/l]"]]))
clin$troponine_max <- as.numeric(gsub(" .*", "", clin[["Troponin -T (max) [ng/l]"]]))
clin$tropo     <- c("NO", "YES")[ (as.numeric(gsub(" .*", "", clin[["Troponin -T [ng/l]"]])) > 14) + 1 ]
clin$tropo_max <- c("NO", "YES")[ (as.numeric(gsub(" .*", "", clin[["Troponin -T (max) [ng/l]"]])) > 14) + 1 ]
message(class(clin$icu))
```

```{r adding_key}
## Add Patient_ID key
key <- read_xlsx("meta_data/January_MK-V7_311020_ps_14.00.xlsx") %>% 
  rename_with(~ gsub("-", "_", .)) %>%
  mutate(Patient_ID=ifelse(Patient_ID == "NA", gsub("  *", "-", gsub(",.*", "", CC_ID)), Patient_ID)) %>%
  mutate(CC_ID=gsub("  *", "", CC_ID)) %>%
  separate_rows(CC_ID, sep=",") %>%
  pivot_longer(starts_with("Sample"), names_to="tube_sample", values_to="SampleID") %>% 
  filter(!(duplicated(Patient_ID) & is.na(SampleID))) %>%
  select(Patient_ID, CC_ID, SampleID, Comments)

clin$Patient_ID <- key$Patient_ID[ match(tolower(clin$ID), tolower(key$CC_ID)) ]
clin$SampleID  <- map_chr(clin$Patient_ID, ~ paste(key$SampleID[ key$Patient_ID == . ], collapse=","))

```

```{r remove_nonconsent}
## remove individuals who retracted their consent
## also, remove individual with no Covid

dropouts <- read_xlsx("../meta_data/Dropout_please delete_PS_V2_7112020.xlsx")
clin <- clin %>% filter(!Patient_ID %in% dropouts[["Pat ID"]])

keys_new <- read_xlsx("../meta_data/Key131020.xlsx", sheet=1) %>%
  select(1:6) %>%
  set_names(c("TubeID", "Comments", "SIDNum", "BHID", "SID", "NEG")) %>%
  mutate(NEG=ifelse(is.na(NEG), "no", NEG)) %>%
  mutate(SIDNum=as.numeric(SIDNum))

covid_neg <- keys_new %>% filter(NEG == "yes")

clin <- clin %>% filter(!(Patient_ID %in% covid_neg$SID | ID %in% covid_neg$BHID))
```

```{r who_icu, cache=TRUE}
## Adding WHO ICU score. Need three additional files for that

## first, the who score
who <- xmlParse("meta_data/Export Demografie, Risikofaktoren und WHO Scale/WHOScale_0901.xml") %>% xmlToDataFrame %>%
  mutate(PatIDNumeric=gsub(".*-0*([1-9]+[0-9]*)$", "\\1", PatID))

who.max    <- who %>% group_by(PatID) %>% summarise(max_whoscore=max(WHOScaleWert, na.rm=TRUE))
who.screen <- who %>% filter(grepl("(Screening)", mnpvislabel))

clin$who     <- who.screen$WHOScaleWert[ match(clin$Patient_ID, who.screen$PatID) ] %>% as.numeric
clin$who.max <- who.max$max_whoscore[ match(clin$Patient_ID, who.max$PatID) ] %>% as.numeric
```

```{r mca_score, cache=TRUE}
foo <- clin %>% select(ID, death, icu, intub, troponine) %>% 
    mutate_all(~ factor(.)) %>% 
    mutate(troponine=as.numeric(as.character(troponine))) %>%
    mutate(troponine=factor(findInterval(troponine, seq(0, 50, by=10))))
    
    as.data.frame
rownames(foo) <- foo$ID
foo$ID <- NULL
mca <- MCA(foo, graph=F)

ind <- get_mca_ind(mca)
clin$MCA_big <- ind$coord[,1]

foo <- clin %>% select(ID, death, icu, intub) %>% 
    mutate_all(~ factor(.)) %>% as.data.frame
rownames(foo) <- foo$ID
foo$ID <- NULL
mca <- MCA(foo, graph=F)

ind <- get_mca_ind(mca)
clin$MCA_small <- ind$coord[,1]
```

```{r clinical_data_spain_switzerland}
#clin_es <- read_xlsx("../meta_data/Data base July2020 (3)_Sevilla_LMR.xlsx") %>%
clin_es <- read_xlsx("../meta_data/Data base July2020_3.xlsx") %>%
  mutate(ID=paste0("ES", .[["COD ext"]])) %>%
  rename(ICU="Admision in Intensive care unit", intub="Intubated") %>%
  replace_na(list(ICU="No")) %>% mutate(ICU=toupper(ICU), intub=toupper(intub)) %>%
  mutate(Severity=c("Mild airway symptoms"="low", "Pneumonia"="high", "asymptomatic"="low", "Severe pneumonia"="high")[ .[["Clinical symptons"]] ])

clin_ch <- read_xls("../meta_data/ETHKSBCovid19Myokard_DATA_1.Ladung.22.07.20.BadenDP.xls") %>%
  mutate(ID=paste0("SB", pid)) %>%
rename(ICU=icu_yn, intub="invasive_ventilation") %>%
  mutate(ICU=c("NO", "YES")[ICU + 1]) %>%
  mutate(intub=c("NO", "YES")[intub + 1])

clin_ch_sum <- clin_ch %>% summary_colorDF
clin_ch <- clin_ch[ , clin_ch_sum$NAs < nrow(clin_ch) & clin_ch_sum$unique > 1 ]

clin_ch2 <- read_xls("../meta_data/covidmyo labor nachbestimmungen 280820_no ID.xls") %>% 
  filter(grepl("^SB", Patientenname))

clin_ch2_w <- clin_ch2 %>% pivot_wider(id_cols=Patientenname, values_from="Resultat (S2)", names_from=Ausdruck) %>% 
  mutate_at(2:5, ~ as.numeric(gsub(" *", "", gsub("^[><]", "", gsub(",", ".", .))))) %>% 
  set_names(c("ID", "TropoT", "CPK", "BNP", "TSH"))

#clin_ch$troponine <- as.numeric(gsub(",", ".", clin_ch2[["Resultat (S2)"]]))[ match(clin_ch$ID, clin_ch2$Patientenname) ]
clin_ch <- merge(clin_ch, clin_ch2_w, by="ID", all.x=TRUE)

clin_eg <- read_xlsx("../meta_data/clinicaldata_cdna_most relevant_FGG.xlsx") %>%
  rename(ID="Sample ID") %>% 
  rename(ICU="ICU Admittance") %>%
  filter(!is.na(ID)) %>%
  mutate(ID=paste0("EG", ID))
clin_eg$ICU[nrow(clin_eg)] <- "NO"
clin_eg$troponine <- as.numeric(gsub(",", ".", gsub("<", "", clin_eg[["Troponin -I pg7ml"]])))
clin_eg$crp <- as.numeric(gsub(",", ".", clin_eg[["CRP (C-reactive protein) [mg/l]"]]))
```

```{r batch2_reading}
hla_new <- hla_new_orig_1a <- read_xlsx("meta_data/2020-08-26_Heidecker_HLA_Typing.xlsx", sheet=1) %>%
  select(ID="Sample name", locus=Locus, alleles="Typing result") %>%
  mutate(Patient_ID=key$Patient_ID[ match(tolower(ID), tolower(key$CC_ID)) ]) %>%
  filter(!Patient_ID %in% patient_IDs_to_exclude)

hla_batch2a <- read_delim("meta_data/Heidecker-Covid-19-Batch2-qm_2020-09-28.csv", delim=";")
hla_batch2b <- read_delim("meta_data/HLA Reads 3-qm_2020-10-07.csv", delim=";")
hla_batch2_full <- rbind(hla_batch2b, hla_batch2a) %>% 
  select(ID="Sample name", locus=Locus, alleles="Typing result") %>%
  filter(!duplicated(paste0(ID, locus)))

## we need to remap the berlin sample IDs
hla_batch2_berlin <- hla_batch2_full %>% filter(grepl("^LV", ID)) %>%
  mutate(Patient_ID=key$Patient_ID[ match(ID, key$SampleID) ]) %>%
  filter(!Patient_ID %in% patient_IDs_to_exclude)

hla_new <- rbind(hla_batch2_berlin, hla_new) %>%
  filter(!duplicated(paste0(Patient_ID, locus)))

#source("problematic_samples.R")
#ids_exclude <- problem_ids %>% filter(!grepl("^OK:", comment)) %>% pull(ids2) %>% strsplit(",") %>% unlist
#hla_new <- hla_new %>% filter(!Patient_ID %in% ids_exclude)

hla_batch2 <- hla_batch2_full %>% filter(!grepl("^LV", ID))
```

```{r coding_alleles,eval=TRUE}
## coding alleles for collaborators, to hide the genetic information from the patients
hla_2code <- hla_new %>% select(CC_ID=ID, Patient_ID, locus, alleles) %>%
  mutate(alleles=ifelse(alleles == "I n s u f f i c i e n t   d a t a", NA, alleles)) %>%
  mutate(alleles_l=strsplit(alleles, ", *")) %>%
  mutate(allele1=map_chr(strsplit(alleles, ", "), ~ .[[1]]),
         allele2=map_chr(strsplit(alleles, ", "), ~ if(length(.) < 2) { .[[1]] } else { .[[2]]} )) %>%
  mutate(allele1=gsub("^.*\\*", "", allele1)) %>% 
  mutate(allele2=gsub("^.*\\*", "", allele2)) %>%
  select(CC_ID, Patient_ID, locus, allele1, allele2) %>% drop_na()

set.seed(20201001)

alleles <- unique(sort(c(hla_2code$allele1, hla_2code$allele2)))
alleles_split <- strsplit(alleles, ":")
names(alleles_split) <- alleles

let_code <- expand.grid(LETTERS, LETTERS) %>% apply(1, paste, collapse="")

recode <- lapply(1:4, function(i) {
  pos <- map_chr(alleles_split, ~ .x[i]) %>% unique() %>% sort
  pos_enc <- sample(let_code, length(pos))
  names(pos_enc) <- pos
  pos_enc
})

encoded <- lapply(1:4, function(i) {
  pos <- map_chr(alleles_split, ~ .x[i]) 
  ret <- recode[[i]][ pos ]
  ret[is.na(ret)] <- ""
  names(ret) <- alleles
  ret
})

encoded <- Reduce(paste, encoded) %>% { gsub("  *$", "", .) } %>% { gsub(" ", ":", .) }
names(encoded) <- alleles

hla_encoded <- hla_2code %>% mutate(
  allele1_encoded=encoded[ allele1 ],
  allele2_encoded=encoded[ allele2 ])

encoding_key <- data.frame(allele=alleles, encoded=encoded)

write_xlsx(list("Unblinded data"=hla_encoded, "Encoding key"=encoding_key), path="hla_set1_unblinded.xlsx")
write_xlsx(hla_encoded %>% select(CC_ID, Patient_ID, locus, allele1_encoded, allele2_encoded), path="hla_set1_blinded.xlsx")
```

```{r viral_load}
viral_load_terry <- read_xlsx("meta_data/viral_load_20201123-patient-study-ids-result.xlsx")
viral_load_guido <- read_csv("meta_data/TC_model_parameters_by_participant.csv") %>% filter(variable == "intercept")
viral_load <- merge(viral_load_terry, viral_load_guido, all.x=TRUE, by.x="Person hash", by.y="personHash") %>%
  filter(!is.na(.data[["First PCR viral load"]])) %>%
  rename(Patient_ID=StudyId, "est_peak_bayes"=mean, "pcr_measurement_01"="First PCR viral load", "est_peak_SA"="Est. peak viral load (SA)") %>%
  select(-variable, -"Person hash")
```

```{r clean_up_hla}
## filter by availability of clinical data
hla_new <- hla_new[ hla_new$Patient_ID %in% clin$Patient_ID, ]

hla <- hla_new %>% select(Patient_ID, locus, alleles) %>%
  mutate(alleles=ifelse(alleles == "I n s u f f i c i e n t   d a t a", NA, alleles)) %>%
  mutate(alleles_l=strsplit(alleles, ", *")) %>%
  mutate(allele1=map_chr(strsplit(alleles, ", "), ~ .[[1]]),
         allele2=map_chr(strsplit(alleles, ", "), ~ if(length(.) < 2) { .[[1]] } else { .[[2]]} )) %>%
  mutate(allele1=gsub("^.*\\*", "", allele1)) %>%
  mutate(allele2=gsub("^.*\\*", "", allele2)) %>%
  mutate(allele1_trunc=gsub("^([0-9]{2}:[0-9]{2}).*", "\\1", allele1)) %>%
  mutate(allele2_trunc=gsub("^([0-9]{2}:[0-9]{2}).*", "\\1", allele2))

hla_l <- hla %>% mutate(homozygous=allele1 == allele2) %>%
         select(Patient_ID, locus, homozygous, allele1, allele2) %>%
         pivot_longer(cols=c(allele1, allele2), names_to="allele_n", values_to="allele_v") %>%
         mutate(allele_trunc=map_chr(allele_v, ~ gsub("^([0-9]{2}:[0-9]{2}).*", "\\1", .))) %>%
         mutate(allele_full=paste0(locus, "*", allele_v)) %>%
         mutate(allele_trunc_full=paste0(locus, "*", allele_trunc))
```

```{r batch2_cleanup}
hla_b2 <- hla_batch2 %>% select(ID, locus, alleles) %>%
  mutate(alleles=ifelse(alleles == "I n s u f f i c i e n t   d a t a", NA, alleles)) %>%
  mutate(alleles_l=strsplit(alleles, ", *")) %>%
  mutate(allele1=map_chr(strsplit(alleles, ", "), ~ .[[1]]),
         allele2=map_chr(strsplit(alleles, ", "), ~ if(length(.) < 2) { .[[1]] } else { .[[2]]} )) %>%
  mutate(allele1=gsub("^.*\\*", "", allele1)) %>%
  mutate(allele2=gsub("^.*\\*", "", allele2)) %>%
  mutate(allele1_trunc=gsub("^([0-9]{2}:[0-9]{2}).*", "\\1", allele1)) %>%
  mutate(allele2_trunc=gsub("^([0-9]{2}:[0-9]{2}).*", "\\1", allele2))

hla_b2 <- hla_b2 %>%
  mutate(site=ifelse(ID %in% clin_ch$ID, "CH", ifelse(ID %in% clin_es$ID, "ES", ifelse(ID %in% clin_eg$ID, "EG", "Unknown"))))

hla_l_b2 <- hla_b2 %>% mutate(homozygous=allele1 == allele2) %>%
         select(ID, locus, homozygous, allele1, allele2) %>%
         pivot_longer(cols=c(allele1, allele2), names_to="allele_n", values_to="allele_v") %>%
         mutate(allele_trunc=map_chr(allele_v, ~ gsub("^([0-9]{2}:[0-9]{2}).*", "\\1", .))) %>%
         mutate(allele_full=paste0(locus, "*", allele_v)) %>%
         mutate(allele_trunc_full=paste0(locus, "*", allele_trunc))
```

```{r clean_up_patients}
patients <- data.frame(Patient_ID=unique(hla$Patient_ID), stringsAsFactors=FALSE)
patients <- merge(patients, clin  %>% 
  mutate(BMI=as.numeric(BMI), weight=as.numeric(.[["Weight [kg]"]])) %>% 
  mutate(CRP=as.numeric(gsub(",", ".", .data[["CRP (C-reactive protein) [mg/l]"]]))) %>%
  select(Patient_ID, icu, intub, troponine, troponine_max, tropo, tropo_max, who, who.max,
  MCA_big, MCA_small,
  age=Age,
  sex=Sex,
  weight, BMI, CRP
  )  %>% filter(!duplicated(Patient_ID)) %>% mutate(age=as.numeric(age))
  , by="Patient_ID")
patients_b1 <- patients
write_tsv(patients, path="patients_dupa")

set.seed(20201023)
train <- sample(patients$Patient_ID, 2/3 * length(patients$Patient_ID))
test <- setdiff(patients$Patient_ID, train)
```

```{r ds1_association,cache=TRUE}
traits <- c("icu", "intub", "tropo", "troponine", "who", "MCA_big")#, "MCA_small")
res_all   <- run_assoc_test_new(hla, traits, patients=patients_b1, id="Patient_ID")
res_train <- run_assoc_test_new(hla %>% filter(Patient_ID %in% train), traits, patients=patients_b1, id="Patient_ID")
res_test  <- run_assoc_test_new(hla %>% filter(Patient_ID %in% test), traits, patients=patients_b1, id="Patient_ID")
hla_all   <- merge(hla, patients, by="Patient_ID")


res_all_summary   <- get_assoc_summary_new(res_all, min_freq, min_patient)
res_train_summary <- get_assoc_summary_new(res_train, min_freq, min_patient)
res_test_summary  <- get_assoc_summary_new(res_test, min_freq, min_patient)

sex_traits <- c("sex", "age", "weight", "BMI")
res_sex <- run_assoc_test_new(hla, sex_traits, patients=patients_b1, id="Patient_ID", log=FALSE)
res_sex_summary <- get_assoc_summary_new(res_sex, min_freq, min_patient)
```

```{r ds1_viral_load_association}
traits <- c("First PCR viral load", "Est. peak viral load (SA)", "Est. Peak (Bayes)")
traits <- c("pcr_measurement_01", "est_peak_bayes")
res_vl <- run_assoc_test_new(hla, traits, patients=viral_load, id="Patient_ID", allele1="allele1_trunc", allele2="allele2_trunc", log=FALSE)
res_vl_summary <- get_assoc_summary_new(res_vl, min_freq, min_patient)
res_all <- c(res_all, res_vl)
res_all_summary   <- get_assoc_summary_new(res_all, min_freq, min_patient)

patients_b1 <- merge(patients_b1, viral_load %>% select("Patient_ID", pcr_measurement_01, est_peak_bayes, est_peak_SA), by="Patient_ID", all.x=TRUE)
```

```{r ds2_associations,cache=TRUE}
res_b2_ch <- run_assoc_test_new(hla_b2, c("ICU", "TropoT", "intub"), patients=clin_ch, id="ID")
res_b2_ch_summary <- get_assoc_summary_new(res_b2_ch, min_freq, min_patient)

res_b2_es <- run_assoc_test_new(hla_b2, c("Hospitalization", "ICU", "intub", "Severity", "sevPn"),
  patients=clin_es %>% mutate(sevPn=ifelse(.data[["Clinical symptons"]] == "Severe pneumonia", "YES", "NO")))
res_b2_es_summary <- get_assoc_summary_new(res_b2_es, min_freq, min_patient)
```





# Notes

In this report version, following `Patient_ID`s were excluded: 

`r patient_IDs_to_exclude %>% paste(collapse=", ")`

Furthermore, all Granada samples were also removed.

In this version of the report, the viral loads have been added to be used
for candidate selection in data set 1. That is, the significant (at p <
  0.05) associations between viral load and HLA were used to include
HLA alleles for validation.

# Methods

## Association test

For testing the association of the HLA data with the outcomes, we have used
the package HIBAG [@zheng2014hibag] with a dominant model. Alleles with a frequency 
lower than `r min_freq <- 0.05;min_freq` or present in fewer than `r min_patient <- 5;min_patient`
were ignored.

Following outcomes were tested: 

 * model: `r model <- "dominant" ; model`
 * association with:
    * ICU 
    * intubation
    * death
    * troponine levels (both as continuous and threshold > 14 ng/l)
    * WHO scores (both as the WHO score at admittance and maximum WHO score over time)
    * composite score: ICU + intubation + death + troponine ("`MCA_big`")
    * composite score: ICU + intubation + death ("`MCA_small`")
 * Data sets:
    * data set 1 (DS1): Berlin cohort
    * data set 2 (DS2): Swiss and Spanish cohorts
    * data set 3 (DS3): RNA-Seq based data set


# Response variables

## Troponine levels vs other predictors

We found that troponine levels differ significantly between groups of
patients defined by either of ICU, intubation status or WHO score (p.value
< 0.001 in one-way ANOVA with log(troponine level) as response).

```{r fig.width=10}
par(mfrow=c(1,3))
beeswarm(troponine ~ icu, data=patients, log=TRUE, pch=19, col="#666666")
beeswarm(troponine ~ intub, data=patients, log=TRUE, pch=19, col="#666666")
beeswarm(troponine ~ who, data=patients, log=TRUE, pch=19, col="#666666")
```

**Table.** Results of one-way ANOVA for the three outcomes (ICU, intubation
status and WHO score) tested.

```{r results="markdown"}
set <- c("icu", "intub", "who")
ret <- map_dfr(set, ~ aov(log(patients$troponine) ~ factor(patients[[.x]])) %>% broom::tidy() %>% slice(1) %>% mutate(term=.x))
pander(ret, digits=2)
```

## ICU score vs other predictors

ICU score was significantly associated with other outcomes (binary
troponine levels, ICU, intubation status), $\chi^2$ test at p < 0.05.


```{r fig.width=10}
par(mfrow=c(1,3))
beeswarm(who ~ tropo, data=patients, pch=19, col="#666666")
beeswarm(who ~ icu, data=patients, pch=19, col="#666666")
beeswarm(who ~ intub, data=patients, pch=19, col="#666666")
```

**Table.** Results of the $\chi^2$ test with WHO score as response
variable.

```{r results="markdown"}
set <- c("tropo", "intub", "who")
map_dfr(set, ~ chisq.test(table(patients$who, patients[[.x]])) %>% broom::tidy() %>% slice(1) %>% mutate(term=.x)) %>%
pander(digits=2)
```

## MCA score (big)

```{r fig.width=10}
par(mfrow=c(1,3))
beeswarm(MCA_big ~ tropo, data=patients, pch=19, col="#666666")
beeswarm(MCA_big ~ icu, data=patients, pch=19, col="#666666")
beeswarm(MCA_big ~ intub, data=patients, pch=19, col="#666666")
```


## MCA score (small)

```{r fig.width=10}
par(mfrow=c(1,3))
beeswarm(MCA_small ~ tropo, data=patients, pch=19, col="#666666")
beeswarm(MCA_small ~ icu, data=patients, pch=19, col="#666666")
beeswarm(MCA_small ~ intub, data=patients, pch=19, col="#666666")
```




## Correlation troponine / troponine max

```{r}
plot(patients$troponine, patients$troponine_max,
  xlab="Troponine",
  ylab="Max. troponine",
  log="xy",
  bty="n")
```

# HLA associations in Data Set 1

## Data set overview

**Table.** Summary overview of the data set 1 patients. 

```{r, results="markdown"}
ds1_full_patient_df <- hla_all %>% 
  select(Patient_ID, locus, alleles, icu:BMI) %>% 
  pivot_wider(id_cols=Patient_ID, names_from=locus, values_from=alleles) %>% { right_join(patients_b1, ., by="Patient_ID") }

write_xlsx(ds1_full_patient_df, path=paste0("full_patient_list_", Sys.Date(), ".xlsx"))

tmp <- ds1_full_patient_df %>% select(Patient_ID:BMI) %>% summary_colorDF %>%
  rename("Missing"=NAs, "Unique values"=unique, Variable=Col, "Variable type"=Class) 
write_xlsx(tmp, path=paste0("ds1_summary_", Sys.Date(), ".xlsx"))
tmp %>% pander(split.tables=Inf)
```


```{r summaries}
by.j <- c("allele", "parameter")
res_summary <- full_join(res_train_summary, res_test_summary, by=by.j, suffix=c("train", "test"))
res_summary <- full_join(res_summary, res_all_summary, by=by.j, suffix=c("", "all"))
```

## Results summary

**Table.** Results of the HLA association testing for all response variables in full DS1 at p~adj~ < 0.5. 

```{r, results="markdown"}
res_all_sign <- res_all_summary %>% filter(p.val < 0.05 & !is.na(p.adj)) %>% 
  filter(!grepl("(peak|pcr|viral load|Peak)", parameter)) %>%
  arrange(p.val)
res_all_sign %>% filter(p.adj < 0.05) %>%
  select(allele, parameter, N=.data[["[-/h,h/h]"]], p.val, p.adj) %>% arrange(p.adj) %>% 
  pander(digits=2, split.tables=Inf)
res_all_sign %>% select(allele, parameter, N=.data[["[-/h,h/h]"]], p.val, p.adj) %>%
  write_xlsx(path="../manuscripts/supplementary_table_candidates.xlsx")



```

## Individual results

```{r}
dt_show <- function(x, param) {
  x %>% 
  filter(freqtrain >= min_freq) %>%
  filter(parameter == param) %>%
  filter(!is.na(p.adjtrain)) %>%
  #mutate("confirmed"=ifelse(!is.na(p.valtrain) & !is.na(p.valtest) & p.valtrain < 0.05 & p.valtest < 0.05, "Confirmed", "")) %>%
  arrange(p.valtrain) %>%
  select(Locus=locus, Allele=allele, Frequency=freq, "p (train)"=p.valtrain, "p (test)"=p.valtest, "p (all)"=p.val, "q (all)"=p.adj) %>%
  datatable(class = 'cell-border stripe') %>%
  formatRound(digits=2, columns=c("Frequency", "p (train)", "p (test)", "p (all)", "q (all)"))

}
```


### Table with individual results {.tabset}

**Table.** Results of the HLA association testing. Only alleles with at
least `r min_freq*100`% frequency and present in at least `r min_patient`
individuals in the training test were included.


#### ICU

```{r, results="markdown"}
res_summary %>% dt_show("icu")
```

#### Intubation

```{r, results="markdown"}
res_summary %>% dt_show("intub")
```

#### WHO score

```{r, results="markdown"}
res_summary %>% dt_show("who")
```

#### Troponine


```{r, results="markdown"}
res_summary %>% dt_show("troponine")
```

#### Troponine above threshold

```{r, results="markdown"}
res_summary %>% dt_show("tropo")
```

#### WHO max

```{r, results="markdown"}
res_summary %>% dt_show("who.max")
```

#### MCA big

```{r, results="markdown"}
res_summary %>% dt_show("MCA_big")
```

#### MCA small

```{r, results="markdown"}
res_summary %>% dt_show("MCA_small")
```


### {-}







## Train vs Test

```{r}
p1 <- "p.valtrain"
p2 <- "p.valtest"
```

Spearman correlation between p-values: 
rho=`r cor( res_summary[[p1]], res_summary[[p2]], method="s", use="p") %>% format(digits=2)`.

```{r trainvstest_ttest,fig.width=6,fig.height=6}
tmp <- res_summary %>% arrange(desc(.data[[p1]] * .data[[p2]]))
plot(
  tmp[[p1]],
  tmp[[p2]],
  xlab="Train",
  ylab="Test",
  log="xy",
  bty="n",
  pch=19,
  col="#CCCCCCCC"
)
abline(h=0.05, col="red")
abline(v=0.05, col="red")
text(  
  tmp[[p1]],
  tmp[[p2]],
  paste0(tmp$allele, "\n", tmp$parameter),
  col=ifelse(tmp[[p1]] < .05 & tmp[[p2]] < .05, "red", "grey"),
  cex=.8
  )
```


## Confirmed results

**Table.** Results of the training / test set confirmation of association
analysis in DS1 (Berlin).  All associations significant in at least one of
the tests are shown for the alleles for
which at least one confirmed association was shown.

```{r results="markdown"}
sel_alleles <- res_summary %>% filter(p.valtrain < .05 & p.valtest < .05) %>%
  pull(allele) %>% unique
res_summary %>% filter(allele %in% sel_alleles) %>%
  filter(p.valtrain < .05 | p.valtest < .05 | p.val < .05) %>%
  arrange(allele, p.val) %>%
  select(allele, parameter, "param. type"=type, freq, "[-/-]", "[-/h,h/h]", "T.[-/-]", "T.[-/h,h/h]", 
    "p(test)"=p.valtest,
    "p(train)"=p.valtrain,
    "p(full)"=p.val,
    "p(adj)"=p.adj) %>%
  pander(split.tables=Inf, digits=2)
```



## Allele candidates

**Table.** Following alleles have been selected as candidates for validation.

```{r results="markdown"}
res_summary %>% filter(p.val < .05 & !is.na(p.adj)) %>%
mutate(confirmed=ifelse(!is.na(p.adjtrain) & !is.na(p.adjtest) & !is.na(p.valtrain) & !is.na(p.valtest) & p.valtrain < .05 & p.valtest < .05, "yes", "no")) %>%
  arrange(p.val) %>%
  mutate(
    freq=format(freq, digits=2),
    p.val=format.pval(p.val, digits=2),
    p.adj=format.pval(p.adj, digits=2)) %>%
  select(Allele=allele, Parameter=parameter, Frequency=freq, "[-/-]", "[-/h,h/h]",
    "P.value"=p.val,
    "FDR"=p.adj,
    confirmed) %>% 
    filter(!grepl("(peak|pcr)", Parameter)) %>%
    mutate(Parameter=c(intub="Intubation", icu="ICU", MCA_big="composite",
    tropo="troponine (cat.)", troponine="troponine (cont.)", who="WHO score")[Parameter]) %>%
    pander(split.tables=Inf)
  #datatable(class = 'cell-border stripe') %>%
  #formatRound(digits=2, columns=c("Frequency", "T.[-/-]", "T.[-/h,h/h]", "p(train)", "p(test)", "p(all)", "p(adj)"))
```


**Fig.** Association between alleles and response parameters in DS1.
Each dot represents one allele / response parameter association.
Left, association with categorical response parameter (ICU status,
intubation status and troponine levels raised above 14 ng/l); right,
associations with continuous parameters (troponine level, WHO score and
composite score). Y axis represents the negative logarithm of the p-value
obtained from corresponding test (left: Fisher's exact test; right:
t-test). Colors correspond to the response parameters. Sizes of
dots correspond to the calculated effect size of the associations (left:
Cramer's v, right: Cohen's d). Faint lines correspond to the selection
thresholds (vertical line: frequency selection threshold of `r min_freq`;
horizontal line: p-value threshold of 0.05).

```{r fig.width=15,fig.height=8}
require(cowplot)
require(ggplot2)
require(hrbrthemes)
require(ggrepel)
set.seed(123)
tmp <- res_all_summary %>% 
  filter(type == "categorical" & parameter != "tropo_max") %>% 
  mutate(Parameter=c(intub="Intubation", icu="ICU", tropo="Troponin T hs > 14ng/l")[parameter]) %>%
  mutate(Effect=E, label="") %>% 
  arrange(freq < min_freq, p.val) %>% 
  mutate(label=ifelse(1:n() > 3, "", allele))

g1 <- ggplot(tmp %>% filter(!is.na(p.val)), aes(x=freq, y=-log10(p.val), size=Effect, group=Parameter)) + 
  geom_point(aes(col=Parameter), alpha=0.5) + 
  geom_hline(yintercept=-log10(.05), col="grey") +
  geom_vline(xintercept=min_freq, col="grey") +
  scale_color_ipsum()  + geom_label_repel(aes(label=label), show.legend=FALSE) +
  xlab("Frequency")  + ggtitle("Categorical")

tmp <- res_all_summary %>% 
  filter(type != "categorical" & parameter %in% c("who", "troponine", "MCA_big")) %>% 
  mutate(Effect=abs(E), label="") %>% 
  mutate(Parameter=c(MCA_big="Composite score", troponine="Troponin T hs [ng/l]", who="WHO score")[parameter]) %>%
  arrange(freq < min_freq, p.val) %>% 
  mutate(label=ifelse(1:n() > 3, "", allele))

g2 <- ggplot(tmp %>% filter(!is.na(p.val)), aes(x=freq, y=-log10(p.val), size=Effect, group=Parameter)) + 
  geom_point(aes(col=Parameter), alpha=0.5) + 
  geom_hline(yintercept=-log10(.05), col="grey") +
  geom_vline(xintercept=min_freq, col="grey") +
  scale_color_ipsum()  + geom_label_repel(aes(label=label), point.padding=0, box.padding=2, max.iter=5000, show.legend=FALSE) +
  xlab("Frequency")  + ggtitle("Continuous")
plot_grid(g1, g2)
```



# Validation with Data Set 2

## Characterization of the cohorts in Data Set 2

 * Switzerland:
    * 21 patients, 9 in the ICU
    * troponine levels determined in all but 4 patients
    * CRP, creatinine, creatinine clearance in all but 3 patients
    * procalcitonine in all but 4

 * Granada (REMOVED):
    * 18 patients, 0 in the ICU
    * creatinine, CRP
    * troponine in all but 3 patients

 * Sevilla:
    * 120 patients, 4 in the ICU
    * clinical symptoms: asymptomatic/mild/pneumonia/severe pneumonia
    * sparse additional data


**Fig.** Measured variables in Swiss data set in dependence on the ICU
status.


```{r swiss_tropo,fig.width=11,fig.height=4}
par(mfrow=c(1, 4))
showgene(clin_ch$crp, clin_ch$ICU, ylab="CRP")
showgene(clin_ch$troponint, clin_ch$ICU, ylab="troponint")
showgene(clin_ch$creatinine, clin_ch$ICU, ylab="Creatinine", ylim=c(50, 150))
showgene(clin_ch$procalcitonine, clin_ch$ICU, ylab="Procalcitonine", log=TRUE) 
```


## Data set 2 HLA typing results

```{r batch2_hlatyping_by_cohort}
.icu   <- res_all$icu %>% reduce(rbind)
.intub <- res_all$intub %>% reduce(rbind)
show_ds2 <- function(x, param, pval.thr=1) {
  x %>% filter(!is.na(p.adj) & p.val < pval.thr) %>% arrange(p.val) %>%
  filter(parameter == param) %>%
  mutate(batch1icu   =.icu$p.val[ match(allele, .icu$allele) ]) %>%
  mutate(batch1intub =.intub$p.val[ match(allele, .icu$allele) ]) %>%
  select(allele, Frequency=freq, "DS2 p value (ICU)"=p.val, "DS1 p value (ICU)"=batch1icu, "DS1 p value (Intubation)"=batch1intub) %>%
  pander(split.tables=Inf,digits=2)
}
```

### Table: Associations in the DS2 cohorts {.tabset}

#### Switzerland, ICU

```{r, results="markdown"}
res_b2_ch_summary %>% show_ds2("ICU")
```

#### Switzerland, Troponine


```{r results="markdown"}
res_b2_ch_summary %>% show_ds2("TropoT")
```

#### Sevilla, ICU

```{r, results="markdown"}
res_b2_es_summary %>% show_ds2("ICU")
```

#### Sevilla, Severity

```{r, results="markdown"}
res_b2_es_summary %>% show_ds2("Severity")
```


#### Sevilla, Hospitalization

```{r, results="markdown"}
res_b2_es_summary %>% show_ds2("Hospitalization")
```

#### Sevilla, Severe pneumonia

```{r, results="markdown"}
res_b2_es_summary %>% show_ds2("sevPn")
```


### {-}


## Data set 2 hla typing results: ICU (in all 3 cohorts)

**Table.** Statistically significant (at uncorrected p < 0.05) associations
between ICU and HLA in full Data Set 2.  Only alleles with frequency above 
`r min_freq` and in at least `r min_patient` were considered.

```{r batch2_hlatyping}
clin_batch2 <- rbind(clin_ch %>% select(ID, ICU, intub, troponine=TropoT), clin_es %>% select(ID, ICU, intub) %>% mutate(troponine=NA))
hla_batch2  <- merge(hla_b2, clin_batch2, by="ID")
res_batch2  <- run_assoc_test_new(hla_batch2, c("ICU", "intub", "troponine"), patients=NULL, allele1="allele1_trunc", allele2="allele2_trunc", id="ID")
res_batch2_summary <- get_assoc_summary_new(res_batch2, min_freq, min_patient)
```


```{r, results="markdown"}
res_batch2_summary %>% 
  filter(!is.na(p.adj) & p.val < 0.05) %>%
  arrange(p.val) %>%
  select(allele, 2:5, "DS2 p value"=p.val) %>%
  pander(split.tables=Inf,digits=2)
```










# Validation with Data Set 3


```{r}
gse_covar <- read.table("meta_data/GSE157103_covariate_file.txt") %>% 
             rename(ID=label, intub=mechanical_ventilation) %>% 
             select(-filename, -md5, -GEO_Accession..exp., -Sample.Name, -associated_controlled.vocabulary_terms, -Experiment, -disease_state)

gse_opti <- read_xlsx("meta_data/s_2020_Heidecker_GSE157103.xlsx", skip=1) %>% select(ID=1, allele="Optitype Hla Type")
gse_opti <- gse_opti %>% separate_rows(allele, sep=";") %>% mutate(locus=gsub("\\*.*", "", allele)) %>% 
  group_by(locus, ID) %>% summarise(allele1=allele[1], allele2=allele[2], locus=locus[1]) %>% arrange(ID) %>%
  merge(gse_covar %>% select(ID, icu, DM, sex, intub, group), by="ID") %>% 
  filter(group == "COVID_19") %>%
  mutate(allele1=gsub(".*\\*", "", allele1), allele2=gsub(".*\\*", "", allele2)) %>%
  arrange(ID, locus)

gse <- read_tsv("../rnaseq_GSE157103_arcas/rnaseq_GSE157103_arcas_alleles_20201111.tsv")



gse_hla <- merge(gse, gse_covar, by="ID") %>% 
  filter(group == "COVID_19") %>%
  mutate(allele1=gsub("^([0-9]+:[0-9]+):[0-9]*$", "\\1", gsub(".*\\*", "", allele1)), allele2=gsub("^([0-9]+:[0-9]+):[0-9]*$", "\\1", gsub(".*\\*", "", allele2))) %>%
  mutate(locus=gsub("^([ABC])$", "HLA_\\1", locus)) %>%
  arrange(ID, locus)


gse_hla_l <- gse_hla %>% select(ID, locus, allele1, allele2) %>%
mutate(homozygous=allele1 == allele2) %>%
pivot_longer(cols=allele1:allele2, names_to="allele_n", values_to="allele") %>% 
mutate(allele_trunc_full=gsub("^([ABC]\\*)", "HLA_\\1", paste0(locus, "*", allele)))
```

```{r arcas_vs_opti}
## there are some annotation differences 
merged <- merge(gse_opti, gse_hla, by=c("ID", "locus")) %>% select(ID, locus, starts_with("allele"))

problem <- merged %>% group_by(ID, locus) %>% summarise(problem=!setequal(c(allele1.x, allele2.x), c(allele1.y, allele2.y))) %>%
  filter(problem)

#merged %>% filter(paste0(ID, locus) %in% paste0(problem$ID, problem$locus))
```


```{r gse_other_vars}
#vars <- c("icu", "intub", "DM", "sex", "fibrinogen", "Charlson_score", "apacheii", "sofa")
vars <- c("icu", "intub")
#gse_loci <- set_names(unique(gse_hla$locus))
gse_loci <- set_names(unique(gsub(" .*", "", res_all_sign$allele)))
gse_res <- run_assoc_test_new(gse_hla, vars, patients=NULL, allele1="allele1", allele2="allele2")
gse_res_summary <- get_assoc_summary_new(gse_res, min_freq, min_patient)
#gse_res_summary <- gse_res_summary %>% filter(allele %in% res_all_sign$allele & !is.na(p.adj))


```

## RNA-Seq results

**Table.** Results of the Data Set 3 (RNA-Seq) association testing significant at
p < 0.05 (Fisher test). Only alleles with a minimal frequency of 
`r min_freq` and present in at least `r min_patient` individuals were
tested.

```{r gse_icu,results="markdown"}
res_b1 <- res_all$icu %>% reduce(rbind)
res_b2 <- res_batch2_summary 

gse_res_summary %>% 
	filter(!is.na(p.adj) & p.val < 0.05) %>% arrange(p.val) %>%
  mutate(pval_b1=res_b1$p.val[ match(allele, res_b1$allele) ]) %>%
  mutate("pval_b1 (intub)"=(res_all$intub %>% reduce(rbind))$p.val[ match(allele, res_b1$allele) ]) %>%
  mutate(pval_b2=res_b2$p.val[ match(allele, res_b2$allele) ]) %>%
  pander(split.tables=Inf, digits=2)
```


# Final analysis with randomization test


```{r randomization,eval=FALSE}
library(BiocParallel)
options(MulticoreParam=MulticoreParam(workers=6))

N <- 1000
traits_b1 <- c("icu", "intub", "tropo", "troponine", "who", "MCA_big")#, "MCA_small")
traits_b2 <- c("ICU", "intub", "troponine")
traits_b3 <- c("icu", "intub")
for(j in 1:10) {
  message(j)
res_rand <- bplapply(1:N, function(i) {

  hla.tmp <- hla
    patients.tmp <- patients_b1
  patients.tmp$Patient_ID <- sample(patients.tmp$Patient_ID)
  res_all.tmp   <- run_assoc_test_new(hla.tmp, traits_b1, patients=patients.tmp, id="Patient_ID")
  res_all_summary.tmp   <- get_assoc_summary_new(res_all.tmp, min_freq, min_patient)
  res_all_sign.tmp <- res_all_summary.tmp %>% filter(p.val < 0.05 & !is.na(p.adj)) %>% arrange(p.val)

  ord <- sample(1:nrow(hla_batch2))
  hla_batch2.tmp <- hla_batch2 %>% mutate_at(all_of(traits_b2), ~ .[ord])
  res_batch2.tmp <- run_assoc_test_new(hla_batch2.tmp, traits_b2, patients=NULL, allele1="allele1_trunc", allele2="allele2_trunc")
  res_batch2_summary.tmp <- get_assoc_summary_new(res_batch2.tmp, min_freq, min_patient) %>%
    filter(!is.na(p.adj)) 

  ord <- sample(1:nrow(gse_hla))
  gse_hla.tmp <- gse_hla %>% mutate_at(all_of(traits_b3), ~ .[ord])
  gse_res.tmp <- run_assoc_test_new(gse_hla.tmp, traits_b3, patients=NULL, allele1="allele1", allele2="allele2")
  gse_res_summary.tmp <- get_assoc_summary_new(gse_res.tmp, min_freq, min_patient) %>% 
    filter(!is.na(p.adj))

  res_confirmed.tmp <- merge(res_batch2_summary.tmp, gse_res_summary.tmp, by=c("locus", "allele"), suffixes=c("_b2", "_b3"))
  res_confirmed.tmp <- merge(res_all_sign.tmp, res_confirmed.tmp) %>%
    filter(p.val_b2 | p.val_b3 < 0.05) %>% 
    select(locus, allele, starts_with("parameter"), starts_with("p.val"))
  return(res_confirmed.tmp)
})

saveRDS(res_rand, file=paste0("res_rand_1000_", j, ".rds"))
}



```









**Table.** Final analysis results. Candidates selected from Data Set 1 were
compared to the association results in the validation Data Sets 2 and 3.
Only alleles with frequency > `r min_freq` and present in more than
`r min_patient` individuals in all three data sets were considered.
P-value was obtained from a randomization testing procedure (replicating
the selection and validation process with randomly matched genotypes and
clinical data). 

```{r final_analysis, results="markdown"}
res_confirmed <- merge(res_batch2_summary %>% filter(allele %in% res_all_sign$allele & !is.na(p.adj)), 
	gse_res_summary %>% filter(allele %in% res_all_sign$allele & !is.na(p.adj)), by="allele", suffixes=c("_b2", "_b3"), all=TRUE)
res_confirmed <- merge(res_all_sign, res_confirmed, by="allele") %>%
    select(locus, allele, starts_with("freq"), starts_with("parameter"), starts_with("p.val")) 
res_rand <- list.files(path=".", pattern="res_rand_1000.*") %>% map(readRDS) %>% reduce(c)
res_final <- res_confirmed %>% 
             group_by(allele) %>% 
             filter(p.val    == min(p.val)) %>% 
             filter(p.val_b2 == min(p.val_b2)) %>% 
             filter(is.na(p.val_b3) | p.val_b3 == min(p.val_b3)) %>% 
             rowwise %>% 
             mutate(p.val_rand=sum(map_lgl(res_rand, ~ 
              any(.x$p.val <= p.val & (is.na(p.val_b3) | is.na(.x$p.val_b3) | .x$p.val_b3 <= p.val_b3) & .x$p.val_b2 <= p.val_b2)))/length(res_rand)) %>% 
             ungroup %>% mutate(p.val_rand_adj=p.adjust(p.val_rand))
res_final %>%
  select(allele, starts_with("freq"), starts_with("p.val")) %>%
  arrange(gsub("HLA", "AAAA", allele)) %>%
      pander(split.tables=Inf, digits=2)
```


**Fig.** Associations of HLA C\*04:01 with categorical response variables.
Each panel shows the difference between patients who were carrying the
allele (HLA C\*04:01 status: Present) and patients who did not carry the allele
(HLA C\*04:01 status: Absent).
Colors correspond to variable status. Vertical axis shows the absolute number of
patients. D1 - DS3: data sets 1 - 3. 

```{r hlac0401_associations, fig.width=15,fig.height=6}

tmp <- rbind(
  res_all_summary %>% mutate(DS="DS1") %>% filter(parameter != "tropo_max"),
  res_batch2_summary %>% mutate(DS="DS2"),
  gse_res_summary %>% mutate(DS="DS3"))  %>%
    filter(type == "categorical" & allele == "HLA_C 04:01") %>%
    mutate(Parameter=c(intub="Intubation", ICU="ICU", icu="ICU", tropo="Troponin T hs > 14ng/l")[parameter]) 
    
    
tmp_long <- tmp %>%
    mutate(Yes_Present  = .data[["T.[-/h,h/h]"]] * .data[["[-/h,h/h]"]],
           Yes_Absent   = .data[["T.[-/-]"]] * .data[["[-/-]"]],
           No_Present  = .data[["[-/h,h/h]"]] - Yes_Present, 
           No_Absent   = .data[["[-/-]"]] - Yes_Absent) %>% 
  pivot_longer(cols=Yes_Present:No_Absent, names_to=c("Parameter Status", "HLA C*04:01 Status"), names_sep="_", values_to="N") %>%
  mutate(id=paste(DS, Parameter, .data[["HLA C*04:01 Status"]]), param=paste(DS, Parameter)) %>%
  mutate("Parameter Status"=paste(Parameter, .data[["Parameter Status"]], sep=": "))

ggplot(tmp_long, aes(x=.data[["HLA C*04:01 Status"]], y=N, fill=.data[["Parameter Status"]])) + geom_bar(stat="identity") + facet_grid(~ param) + theme(strip.background =element_rect(fill="white")) + scale_fill_brewer(palette="Paired")  
```

**Fig.** Associations of HLA C\*04:01 with categorical response variables.
Each panel shows the difference between patients who were carrying the
allele (HLA C\*04:01 status: Present) and patients who did not carry the allele
(HLA C\*04:01 status: Absent).
Colors correspond to variable status. Vertical axis shows the frequencies of
patients. D1 - DS3: data sets 1 - 3. 


```{r hlac0401_associations_2, fig.width=15,fig.height=6}
tmp_long <- tmp %>%
    mutate(Yes_Present  = .data[["T.[-/h,h/h]"]],# * .data[["[-/h,h/h]"]],
           Yes_Absent   = .data[["T.[-/-]"]],# * .data[["[-/-]"]],
           No_Present  = 1 - .data[["T.[-/h,h/h]"]], #.data[["[-/h,h/h]"]] - Pos_Present,
           No_Absent   = 1 - .data[["T.[-/-]"]]#.data[["[-/-]"]] - Pos_Absent
           ) %>%
  pivot_longer(cols=Yes_Present:No_Absent, names_to=c("Parameter Status", "HLA C*04:01 Status"), names_sep="_", values_to="Frequency") %>%
  mutate(id=paste(DS, Parameter, .data[["HLA C*04:01 Status"]]), param=paste(DS, Parameter)) %>%
  mutate("Parameter Status"=paste(Parameter, .data[["Parameter Status"]], sep=": "))

ggplot(tmp_long, aes(x=.data[["HLA C*04:01 Status"]], y=Frequency, fill=.data[["Parameter Status"]])) + geom_bar(stat="identity") + facet_grid(~ param) + theme(strip.background =element_rect(fill="white")) + scale_fill_brewer(palette="Paired")  


```

# Analysis of the joined data set

```{r joined_data_set}
batch1_hla <- hla_all %>% mutate(INTUB=ifelse(intub=="0", "no", "yes")) %>% select(ID=Patient_ID, locus, allele1=allele1_trunc, allele2=allele2_trunc, ICU=icu, INTUB)
batch2_hla <- hla_batch2 %>% select(ID, locus, allele1=allele1_trunc, allele2=allele1_trunc, ICU, INTUB=intub)
batch3_hla <- gse_hla %>% select(ID, locus, allele1, allele2, ICU=icu, INTUB=intub)
batch123_hla <- rbind(batch1_hla, batch2_hla, batch3_hla) %>% mutate(ICU=toupper(ICU), INTUB=toupper(INTUB)) %>% filter(!is.na(ICU))

batch123_stat <- batch123_hla %>% group_by(locus) %>% summarize(n=n())
batch123_loci <- batch123_stat %>% filter(n > 100) %>% pull(locus) %>% set_names()

res123 <- run_assoc_test_new(batch123_hla, c("ICU", "INTUB"), patients=NULL, allele1="allele1", allele2="allele2", id="ID")
res123_summary <- get_assoc_summary_new(res123, min_freq, min_patient)
```

**Table.** Results of the joined analysis of association with ICU status
for all three data sets. Alleles present in fewer than `r min_patient`
individuals or at a frequency lower than `r min_freq` were not tested for
association.

```{r res123_summary_icu, results="markdown"}
res123_summary %>%
	filter(!is.na(p.adj) & p.val < .05) %>%
  filter(parameter == "ICU") %>%
  select(-type) %>%
      arrange(p.val) %>%
      pander(split.tables=Inf, digits=2)

res123_summary %>% 
	filter(!is.na(p.adj) & p.val < .05) %>%
  select(-type) %>% arrange(p.val) %>%
  write_xlsx(path="../manuscripts/supplementary_table_joined_analysis.xlsx")

```

**Table.** Results of the joined analysis of association with intubation status
for data sets 1 and 3. Alleles present in fewer than `r min_patient`
individuals or at a frequency lower than `r min_freq` were not tested for
association.

```{r res123_summary_intub, results="markdown"}
res123_summary %>%
	filter(!is.na(p.adj) & p.val < .05) %>%
  filter(parameter == "INTUB") %>%
  select(-type, -parameter) %>%
      arrange(p.val) %>%
      pander(split.tables=Inf, digits=2)
```

```{r joined_data_set_tropo}
batch1_hla <- hla_all %>% select(ID=Patient_ID, locus, allele1=allele1_trunc, allele2=allele2_trunc, TROPO=troponine) %>%
  drop_na()
batch2_hla <- merge(clin_ch, hla_batch2) %>% mutate(TROPO=TropoT) %>%
  select(ID, locus, allele1=allele1_trunc, allele2=allele1_trunc, TROPO)
batch123_hla <- rbind(batch1_hla, batch2_hla)

batch123_stat <- batch123_hla %>% group_by(locus) %>% summarize(n=n())
batch123_loci <- batch123_stat %>% filter(n > 100) %>% pull(locus) %>% set_names()

res123$TROPO <- run_assoc_test_new(batch123_hla, c("TROPO"), patients=NULL, allele1="allele1", allele2="allele2", id="ID")$TROPO
res123_summary <- get_assoc_summary_new(res123, min_freq, min_patient)
```

**Table.** Results of the joined analysis of association with troponine
levels
for data sets 1 and 2\_ch. Alleles present in fewer than `r min_patient`
individuals or at a frequency lower than `r min_freq` were not tested for
association.

```{r res123_summary_tropo, results="markdown"}
res123_summary %>%
	filter(!is.na(p.adj) & p.val < .05) %>%
  filter(parameter == "TROPO") %>%
  select(-type, -parameter) %>%
      arrange(p.val) %>%
      pander(split.tables=Inf, digits=2)
```

# Association with KIR

# Association of alleles with viral load

```{r}
require(rstatix)
foo <- merge(clin, viral_load, by="Patient_ID")
  #rename(firstPCR="First PCR viral load", estPeakSA="Est. peak viral load (SA)", estPeakBayes="Est. Peak (Bayes)")
wilco.icu    <- map_dfr(c("pcr_measurement_01", "est_peak_bayes"), ~ wilcox_test(foo, as.formula(paste0(.x, " ~ icu"))))
wilco.p.icu <- set_names(wilco.icu %>% pull(p), wilco.icu %>% pull(.y.))
wilco.ef.icu <- map_dfr(c("pcr_measurement_01", "est_peak_bayes"), ~ wilcox_effsize(foo, as.formula(paste0(.x, " ~ icu"))))
wilco    <- map_dfr(c("pcr_measurement_01", "est_peak_bayes"), ~ wilcox_test(foo, as.formula(paste0(.x, " ~ intub"))))
wilco.p <- set_names(wilco %>% pull(p), wilco %>% pull(.y.))
wilco.ef <- map_dfr(c("pcr_measurement_01", "est_peak_bayes"), ~ wilcox_effsize(foo, as.formula(paste0(.x, " ~ intub"))))

corTropoFirst <- cor.test(foo$pcr_measurement_01, foo$troponine, use="p", method="s")
res_vl_sign <- res_vl_summary %>% filter(!is.na(p.adj) & p.val < .05)
```

For DS1, viral loads as well as estimated peak viral load were available.
Notably, the viral loads were not associated with higher severity of
symptoms. The viral loads and estimated peak viral loads were lower
in the intubated patients, albeit not significantly
(p-values from two-sided Wilcoxon test, respectively,
`r format.pval(wilco.p[1], 2)` and
`r format.pval(wilco.p[2], 2)`). 
The estimated peak size was significantly lower in ICU patients
(`r wilco.p.icu[2]`).
The correlation between troponine levels and 
the initial viral load (but not the estimated peak viral load) was
significant and  positive, but small 
(Spearman's $\rho$=`r format(corTropoFirst$estimate, 2)`, 
p=`r format.pval(corTropoFirst$p.value, 2)`). Furthermore, both measures of
viral load were significantly and negatively correlated with the composite
score.

There were 
`r sum(res_vl_sign$p.val < .05)`
significant associations between HLA at p < 0.05 and none significant after correction for false discovery rate.
The HLA
C\*04:01 allele was not significantly associated with any viral load
variable. The alleles significantly associated with one of the viral load
variables were not formally validated with a randomization test by either
data set 2 or 3 given that higher viral load does not correlate with higher
severity, but none of the alleles was significantly associated with any of
the other tested covariates (such as ICU or intubation status)in data sets
1-3.

**Table XX.** Significant associations between viral load and HLA alleles
in data set 1. 
Allele, two-digit allele code; covariate, the covariate for which the
association was tested; Frequency: allele frequency; [-/-]: number of  
individuals not carrying the allele; [-/h,h/h]: number of individuals
carrying the allele; Effect size: Cohen's d; 
P value: p-value from t-test; FDR: p-value corrected for false discovery
rate.

```{r vl_associations, results="markdown"}
res_vl_sign %>%
  arrange(p.val) %>%
  #filter(parameter != "Est. Peak (Bayes)") %>%
  filter(parameter != "est_peak_SA") %>%
  mutate(parameter=gsub(" \\(SA\\)", "", parameter)) %>%
  mutate(parameter=c(pcr_measurement_01="First measurement",
  est_peak_bayes="Estimated peak viral load")[parameter]) %>%
  select(Allele=allele, Covariate=parameter,
    Frequency=freq,
    "[-/-]", "[-/h,h/h]",
    "Effect size"=E,
    "P value"=p.val,
    "FDR"=p.adj) %>%
  pander(digits=2, split.tables=Inf)
```

**Fig.** Viral load at first measurement was not associated significantly
with ICU / intubation status or troponine levels.

```{r viral_load_overview,fig.width=10,fig.height=4}
vl_init <- "pcr_measurement_01"
vl_sa   <- "est_peak_SA"
vl_bayes <- "est_peak_bayes"

vl_vars <- c(vl_init, vl_sa, vl_bayes)
pr_vars <- c("icu", "intub", "troponine")
foo <- merge(clin, viral_load, by="Patient_ID")

#par(mfrow=c(3,3)) 
par(mfrow=c(1,3)) 
prnam <- c(icu="ICU status", intub="Intubation status")
foo$intub <- c("0"="NO", "1"="YES")[foo$intub]

#for(vl in vl_vars) {
  vl <- vl_init
  for(pr in pr_vars) {
    if(pr == "troponine") {
      plot(foo[[pr]], foo[[vl]], pch=19, col="grey", log="x", bty="n", xlab="Troponin T hs (ng/l)", ylab="Viral load")
    } else {
      comp <- paste(unique(foo[[pr]]), collapse="-")
      showgene(foo[[vl]], foo[[pr]], col="grey", pch=19, main=prnam[pr], ylab="Viral load", comparisons=comp)#, test.func=t.test)
    }
  }
#}
```



# Comparison of frequencies

## Batch1 vs Batch2

```{r fig.width=11,fig.height=5}
sel <- grepl("^LV", hla_l_b2$ID)
hla_berlin <- rbind(hla_l, hla_l_b2 %>% filter(grepl("^LV", ID))) %>% group_by(allele_trunc_full) %>% summarise(locus=locus[1], n=n()) %>% group_by(locus) %>% mutate(frequency=n/sum(n))
hla_spain <- hla_l_b2 %>% filter(grepl("^E[SG]", ID)) %>% group_by(allele_trunc_full) %>% summarise(locus=locus[1], n=n()) %>% group_by(locus) %>% mutate(frequency=n/sum(n))
hla_swiss <- hla_l_b2 %>% filter(grepl("^SB", ID)) %>% group_by(allele_trunc_full) %>% summarise(locus=locus[1], n=n()) %>% group_by(locus) %>% mutate(frequency=n/sum(n))
hla_gse   <- gse_hla_l %>% group_by(allele_trunc_full) %>% summarise(locus=locus[1], n=n()) %>% group_by(locus) %>% mutate(frequency=n/sum(n))

berlin_vs_spain <- merge(hla_berlin, hla_spain, by="allele_trunc_full", all=TRUE) %>% mutate(col=ifelse(is.na(frequency.x) | is.na(frequency.y), "grey", "darkblue")) %>%
  replace_na(list(frequency.x=0, frequency.y=0))
berlin_vs_swiss <- merge(hla_berlin, hla_swiss, by="allele_trunc_full", all=TRUE) %>% mutate(col=ifelse(is.na(frequency.x) | is.na(frequency.y), "grey", "darkblue")) %>%
  replace_na(list(frequency.x=0, frequency.y=0))
berlin_vs_gse <- merge(hla_berlin, hla_gse, by="allele_trunc_full", all=TRUE) %>% mutate(col=ifelse(is.na(frequency.x) | is.na(frequency.y), "grey", "darkblue")) %>%
  replace_na(list(frequency.x=0, frequency.y=0))


par(mfrow=c(1,3)) 

with(berlin_vs_spain, plot(frequency.x, frequency.y, pch=19, col=col, xlab="Berlin", ylab="Spain", bty="n"))
abline(0, 1, col="grey")
with(berlin_vs_swiss, plot(frequency.x, frequency.y, pch=19, col=col, xlab="Berlin", ylab="Switzerland", bty="n")) 
abline(0, 1, col="grey")
with(berlin_vs_gse, plot(frequency.x, frequency.y, pch=19, col=col, xlab="Berlin", ylab="GSE157103", bty="n")) 
abline(0, 1, col="grey")
```

## Comparison of frequencies

```{r fig.width=11,fig.height=5}
pop8 <- read_tsv("../meta_data/allele_frequencies_population_8.txt") %>% rename(frequency="Allele Frequency")
spain <- read_tsv("../meta_data/allele_frequencies_spain.txt") %>% rename(frequency="Allele Frequency")
spain_vs_pop8 <- merge(spain, pop8, by="Allele", all=TRUE) %>% 
  mutate(col=ifelse(is.na(frequency.x) | is.na(frequency.y), "grey", "darkblue")) %>% replace_na(list(frequency.x=0, frequency.y=0))
pop8_vs_berlin <- merge(hla_berlin %>% mutate(allele_trunc_full=gsub("HLA_", "", allele_trunc_full)), pop8, by.x="allele_trunc_full", by.y="Allele", all=TRUE) %>% 
  mutate(col=ifelse(is.na(frequency.x) | is.na(frequency.y), "grey", "darkblue")) %>% replace_na(list(frequency.x=0, frequency.y=0))
pop8_vs_gse <- merge(hla_gse %>% mutate(allele_trunc_full=gsub("HLA_", "", allele_trunc_full)), pop8, by.x="allele_trunc_full", by.y="Allele", all=TRUE) %>% 
  mutate(col=ifelse(is.na(frequency.x) | is.na(frequency.y), "grey", "darkblue")) %>% replace_na(list(frequency.x=0, frequency.y=0))
par(mfrow=c(1,3)) 
with(spain_vs_pop8, plot(frequency.x, frequency.y, pch=19, col=col, ylab="Germany / pop 8", xlab="Spain", bty="n")) 
abline(0, 1, col="grey")
with(pop8_vs_berlin, plot(frequency.y, frequency.x, pch=19, col=col, xlab="Germany / pop 8", ylab="Berlin", bty="n", ylim=c(0, .04), xlim=c(0, .1)))
abline(0, 1, col="grey")
with(pop8_vs_gse, plot(frequency.y, frequency.x, pch=19, col=col, xlab="Germany / pop 8", ylab="GSE157103", bty="n", ylim=c(0, .04), xlim=c(0, .1)))
abline(0, 1, col="grey")
```


## Association between HLAs and other covariates


```{r}
foo <- merge(hla, patients_b1, by="Patient_ID")
ttest.age.icu <- t.test(foo$age ~ foo$icu)
ttest.age.intub <- t.test(foo$age ~ foo$intub)

freq.overall <- foo %>% filter(locus == "HLA_C") %>% mutate(n=(allele1_trunc == "04:01") + (allele2_trunc == "04:01")) %>% mutate(ret=sum(n)/(2*n())) %>% pull(ret) %>% `[`(1)
freq.over75   <- foo %>% filter(age > 75) %>% filter(locus == "HLA_C") %>% mutate(n=(allele1_trunc == "04:01") + (allele2_trunc == "04:01")) %>% 
  mutate(ret=sum(n)/(2*n())) %>% pull(ret) %>% `[`(1)
freq.below65  <- foo %>% filter(age < 65) %>% filter(locus == "HLA_C") %>% mutate(n=(allele1_trunc == "04:01") + (allele2_trunc == "04:01")) %>% 
  mutate(ret=sum(n)/(2*n())) %>% pull(ret) %>% `[`(1)

res_sex_sign <- res_sex_summary %>% arrange(p.val) %>% filter(!is.na(p.adj) & p.val < 0.05)
```

One possibility to explain the association between HLA
alleles and COVID-19 severity is that these alleles occur at different
frequencies in different risk groups. For example, maybe HLA C\*04:01 has a
higher frequency among older patients, who are at greater risk of
developing severe COVID symptoms. Indeed, both patients who were intubated and stationed at
the ICU were significantly older than others (p-values from a two-sided
t-test, respectively,
`r format.pval(ttest.age.intub$p.value, digits=2)`
and
`r format.pval(ttest.age.icu$p.value, digits=2)`). 
However, for the HLA C\*04:01, the opposite was true: while the allele had
an overall frequency of 
`r format(freq.overall, digits=2)`
and 
`r format(freq.below65, digits=2)`
in patients younger than 65, the frequency of the allele was actually lower
in the high-risk group of patients older than 75 years, 
`r format(freq.over75, digits=2)`.

To exclude other potential confounding effects, we have tested the
associations of all alleles with the covariates gender, age, weight and
BMI. While there were  
`r length(unique(res_sex_sign$allele))` 
alleles which showed a significant associations at p < 0.05 (see Supplementary Table
XX), no association was significant after correction for multiple testing.


**Supplementary Table XX.** Associations of HLA alleles and age, gender,
weight and BMI. Only alleles which pass the specified thresholds 
(frequency > `r min_freq` and occurs in at least `r min_patient`)
have been tested for association.

```{r results="markdown"}
res_sex_sign %>%
  mutate(freq=format(freq, digits=2),
         E=format(E, digits=2),
         p.val=format.pval(p.val, digits=2),
         p.adj=format.pval(p.adj, digits=2)) %>%
  select(Allele=allele, Covariate=parameter,
    Frequency=freq,
    "Effect size"=E,
    "P value"=p.val,
    "FDR"=p.adj) %>%
    pander(split.tables=Inf)

```


## Affinity analysis

```{r}
aff2 <- read_xls("meta_data/affinity/nguyen_affinity_appendix_3.xls", skip=1, sheet=1) %>% 
  group_by(Allele) %>% 
  summarise(n_tightly=sum(Binding_affinity < 50), n_loosly=sum(Binding_affinity < 500)) %>% 
  rename(allele=Allele) %>% 
  mutate(allele=gsub("HLA-([ABC])", "HLA_\\1 ", allele))

aff2_orig_50 <- read_xls("meta_data/affinity/nguyen_affinity_appendix_3.xls", skip=1, sheet=1) %>% 
  rename(allele=Allele) %>% 
  mutate(allele=gsub("HLA-([ABC])", "HLA_\\1 ", allele)) %>% 
  filter(Binding_affinity < 50)

aff2_orig_500 <- read_xls("meta_data/affinity/nguyen_affinity_appendix_3.xls", skip=1, sheet=1) %>% 
  rename(allele=Allele) %>% 
  mutate(allele=gsub("HLA-([ABC])", "HLA_\\1 ", allele)) %>% 
  filter(Binding_affinity < 500)



```

A possible explanation for the effect of an allele on disease severity
might be an abnormally high or low binding affinity to the SARS-Cov-2
peptides. To test that hypothesis, we have used a previously described data
set of putative (computed in silico) binding affinities (ref: Nguyen A,
David JK, Maden SK, Wood MA, Weeder BR, Nellore A, Thompson RF. Human
leukocyte antigen susceptibility map for SARS-CoV-2. Journal of virology.
2020 Apr 17.). Following an approach described by (ref: Iturrieta-Zuazo,
Ignacio, et al. "Possible role of HLA class-I genotype in SARS-CoV-2
infection and progression: A pilot study in a cohort of Covid-19 Spanish
patients." Clinical Immunology 219 (2020): 108572.), for each allele we
have calculated the number of SARS-Cov-2 peptides which bind either
"tightly" (at < 50 mM) or "loosely" (at < 500 mM). 

We noticed that HLA C\*04:01 is among ten alleles with the fewest predicted
binding Sars-Cov-2 peptides (Fig. XX). We thus wanted to systematically
test the hypothesis that the putative affinity to SARS-Cov-2 peptides is a
correlate of disease severity. First, for each of the response variables
(ICU and intubation status, troponine levels, composite score) we tested
whether the effect size is correlated to the number of potentially binding
peptides. We included all alleles (not filtered for frequency) in this
comparison. While in all data sets large effects were never seen in alleles
with a large number of potentially binding peptides, the negative
correlation was only significant in DS3 (Spearman test, p < 0.007, $\rho$=0.27). 

Next, for each patient, we have calculated how many SARS-Cov-2 peptides can
type I alleles present in that patient potentially bind and tested whether
there is an association between disease severity (using troponine levels,
intubation and ICU status and composite score as proxies) and that number. However,
none of the tests showed a significant association.


**Fig. XX.** Correlation between the estimated effect size for the
association between alleles and intubation status and the number of
peptides putatively recognized by that allele. Each dot corresponds to one
allele. Left column shows plots for DS1, right column shows plots for DS3;
upper row shows plots for number of "tightly" binding peptides, lower 
row shows plots for number of "loosely" binding peptides. Dot size
corresponds to the p-value in Fisher's exact test. Alleles showing
significant associations with intubation status are labelled.


```{r fig.width=14,fig.height=14}
tmp <- merge(res_all_summary %>% filter(parameter == "intub"), aff2, by="allele") %>% 
  filter(!is.na(p.adj)) %>% mutate(label=ifelse(p.val < 0.05 & !is.na(p.adj), allele, ""))
g1 <- ggplot(tmp,
  aes(x=E, y=n_tightly, size=-log10(p.val))) + 
  geom_point(col="#33333366") + 
  geom_label_repel(aes(label=label), show.legend=FALSE) + ylab("Number of tightly binding peptides")
g3 <- ggplot(tmp,
  aes(x=E, y=n_loosly, size=-log10(p.val))) + 
  geom_point(col="#33333366") + 
  geom_label_repel(aes(label=label), show.legend=FALSE) + ylab("Number of loosely binding peptides")


tmp <- merge(gse_res_summary %>% filter(parameter == "intub"), aff2, by="allele") %>% 
  filter(!is.na(p.adj)) %>% mutate(label=ifelse(p.val < 0.05 & !is.na(p.adj), allele, ""))
g2 <- ggplot(tmp, 
  aes(x=E, y=n_tightly, size=-log10(p.val))) + 
  geom_point(col="#33333366") + 
  geom_label_repel(aes(label=label), show.legend=FALSE) + ylab("Number of tightly binding peptides")
g4 <- ggplot(tmp, 
  aes(x=E, y=n_loosly, size=-log10(p.val))) + 
  geom_point(col="#33333366") + 
  geom_label_repel(aes(label=label), show.legend=FALSE) + ylab("Number of loosely binding peptides")
plot_grid(g1, g2, g3, g4, labels=c("DS1", "DS3", "DS1", "DS3"), nrow=2)
```

```{r}

tmp0 <- hla %>% 
  filter(grepl("HLA_", locus)) %>% 
  select(Patient_ID, locus, a1=allele1_trunc, a2=allele2_trunc) %>% 
  mutate(a1=paste(locus, a1), a2=paste(locus, a2)) %>% 
  pivot_longer(cols=c(a1, a2), names_to="pos", values_to="allele") %>% 
  mutate(n_loosly=aff2$n_loosly[ match(allele, aff2$allele) ], n_tightly=aff2$n_tightly[ match(allele, aff2$allele) ]) %>% 
  group_by(Patient_ID) %>% 
  summarise(n_tightly=sum(n_tightly,na.rm=T), n_loosly=sum(n_loosly, na.rm=T)) %>%
  full_join(patients_b1, by="Patient_ID")

Npept <- function(aff2_orig, alleles) {
  aff2_orig %>% filter(allele %in% alleles) %>%
  filter(!duplicated(Peptide)) %>% nrow
}

tmp <- hla %>% 
  filter(grepl("HLA_", locus)) %>% 
  filter(locus == "HLA_C") %>%
  select(Patient_ID, locus, a1=allele1_trunc, a2=allele2_trunc) %>% 
  mutate(a1=paste(locus, a1), a2=paste(locus, a2)) %>% 
  pivot_longer(cols=c(a1, a2), names_to="pos", values_to="allele") %>% 
  group_by(Patient_ID) %>% 
  summarise(n_tightly=Npept(aff2_orig_50, allele), n_loosely=Npept(aff2_orig_500, allele)) %>%
  full_join(patients_b1, by="Patient_ID")


```




# Population stratification

## Methods

Exome data was used to investigate the genetic stratification in the
analysed samples. For variant calling, XXX[Manuel] was used. The variant
data was then evaluated using the R package SNPRelate, v. 1.20.1.





## Results


```{r fig.width=12,fig.height=6}
library(SNPRelate)
#snpgdsVCF2GDS("meta_data/exome/merged.vcf", "meta_data/exome/merged.gds")
if(exists("genofile")) snpgdsClose(genofile)
genofile <- snpgdsOpen("meta_data/exome/merged.gds")
batch1_hla <- hla_all %>% mutate(INTUB=ifelse(intub=="0", "no", "yes")) %>% select(ID=Patient_ID, locus, allele1=allele1_trunc, allele2=allele2_trunc, ICU=icu, INTUB)
batch2_hla <- hla_batch2 %>% select(ID, locus, allele1=allele1_trunc, allele2=allele1_trunc, ICU, INTUB=intub)
batch12_hla <- rbind(batch1_hla, batch2_hla) %>% mutate(ICU=toupper(ICU), INTUB=toupper(INTUB)) %>% filter(!is.na(ICU))
batch12_hla_w <- batch12_hla %>% pivot_wider(id_cols=c(ID, ICU, INTUB), names_from=locus, values_from=c(allele1, allele2))

snp_ids <- read.gdsn(index.gdsn(genofile, path="sample.id"))
snp_hla <- data.frame(SNP_id=snp_ids) %>%
  mutate(CC_ID=gsub("-N1.*", "", SNP_id)) %>% 
  left_join(key, by="CC_ID") %>% 
  mutate(ID=ifelse(!is.na(Patient_ID), Patient_ID, CC_ID)) %>%
  select(SNP_id, ID) %>% filter(!duplicated(SNP_id)) %>%
  mutate(Cohort=ifelse(grepl("^ES", ID), "ES", ifelse(grepl("^SB", ID), "CH", "DE"))) %>%
  left_join(batch12_hla_w, by="ID") %>% 
  mutate(HLAC0401=ifelse(allele1_HLA_C == "04:01" | allele2_HLA_C == "04:01", "Present", "Absent")) %>%
  filter(!is.na(HLAC0401))

snpset <- snpgdsLDpruning(genofile, ld.threshold=0.2)
snp.pca <- snpgdsPCA(genofile, snp.id=unlist(snpset))
sel <- snp.pca$sample.id %in% snp_hla$SNP_id
pca <- snp.pca$eigenvect[sel, ]

ds1_ids <- intersect(snp_ids[ !grepl("^(SB|ES)", snp_ids) ], snp_hla$SNP_id)
ds2_ids <- intersect(snp_ids[  grepl("^(SB|ES)", snp_ids) ], snp_hla$SNP_id)

snp_pca_ds1 <- snpgdsPCA(genofile, sample.id=ds1_ids, snp.id=unlist(snpset))
sel_ds1 <- match(ds1_ids, snp_hla$SNP_id)

snp_pca_ds2 <- snpgdsPCA(genofile, sample.id=ds2_ids, snp.id=unlist(snpset))
sel_ds2 <- match(ds2_ids, snp_hla$SNP_id)

## check for association with HLA 04:01
## check for association with intub / ICU
## use only DS1


par(mfrow=c(1,2))
plot(
  snp_pca_ds1$eigenvect[,1], 
  snp_pca_ds1$eigenvect[,2], 
  pch=19, col="grey", bty="n", xlab="PC 1", ylab="PC 2", main="Data set 1")
abline(h=0, col="darkgrey", lty=2)
abline(v=0, col="darkgrey", lty=2)

plot(
  snp_pca_ds2$eigenvect[,1], 
  snp_pca_ds2$eigenvect[,2], 
  pch=19, col="grey", bty="n", xlab="PC 1", ylab="PC 2", main="Data set 2")
abline(h=0, col="darkgrey", lty=2)
abline(v=0, col="darkgrey", lty=2)


```



# SessionInfo


```{r results="markdown"}
sessionInfo()
```



